---
title: "Anchor Residuals"
output:
  html_document:
    df_print: paged
    toc: yes
---
Here we demonstrate how to calculate rediduals for an anchor screen.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(messagae = FALSE, warning = FALSE)
```

First we check for installation of packages that we will need for the analysis.
```{r}
list.of.packages <- c("ggplot2", "dplyr", "here", "readr", 'mgcv', 'splines', 
                      'purrr', 'ggrepel')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(here)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```

# Load Data
To calculate residuals need guide-level log-fold changes, a mapping between guides and genes, and a reference for pairs of control arms and treatments. 
```{r}
output_file_root <- 'example'
guide_lfcs <- read_csv(here('guide_lfcs.csv')) # 'here' sets the root to the project's directory
head(guide_lfcs)
guide_gene_mapping <- read_csv(here('guide_gene_mapping.csv'))
head(guide_gene_mapping)
reference_arms <- read_csv(here('reference_arms.csv'))
head(reference_arms)
```

# Calculating Residuals
We merge our dataframes  
```{r}
tidy_lfcs <- guide_lfcs %>%
  gather(treatment, Avg.LFC, -`Construct Barcode`) # Make data columnar
reference_treatment_lfcs <- tidy_lfcs %>%
  inner_join(reference_arms, by = c('treatment' = 'Reference')) %>%
  rename(reference = treatment, treatment = Perturbed) %>%
  inner_join(tidy_lfcs, by = c('treatment','Construct Barcode'), 
             suffix = c('.reference', ''))
head(reference_treatment_lfcs)
```
And then calculate the residuals for each construct in each treatment.
```{r}
guide_residuals <- reference_treatment_lfcs %>%
  group_by(treatment) %>%
  mutate(residual = mgcv::gam(Avg.LFC ~ splines::ns(Avg.LFC.reference, df = 3))$residual)
```

As an example, we can visualize the residuals for the first condition.
```{r}
random_treatment <- sample(guide_residuals$treatment, 1)
ggplot(guide_residuals %>% filter(treatment == random_treatment)) +
  aes(x = Avg.LFC.reference, y = Avg.LFC, color = residual) +
  geom_point(alpha = 0.3) +
  theme_classic() +
  theme(aspect.ratio = 1, text = element_text(size = 12), 
        plot.title = element_text(size = 12)) +
  scale_color_gradient2(low = 'red', mid = 'white', high = 'blue') +
  ggtitle(random_treatment)
```

# Aggregating Residuals 
Here we use the mapping file to aggreagate residuals at the gene level. We Z-score residuals to determine significance levels. 
```{r}
guide_residuals_mapped <- guide_residuals %>%
  inner_join(guide_gene_mapping)
population_stats = guide_residuals_mapped %>% # Stats to be used for z-scoring
  group_by(treatment) %>%
  summarise(pop_mean = mean(residual), 
            pop_sd = sd(residual))
gene_residuals = guide_residuals_mapped %>%
  group_by(treatment) %>%
  nest() %>% # now each treatment is its own nested dataframe
  inner_join(population_stats) %>%
  mutate(gene_residuals = purrr::pmap(list(data, pop_mean, pop_sd), function(df, m, s) {
           df %>%
             group_by(`Gene Symbol`) %>%
             summarise(Avg.Residual = mean(residual),
                       Avg.LFC = mean(Avg.LFC),
                       n_guides = length(residual),
                       z = (Avg.Residual - m)/(s/sqrt(n_guides)), 
                       p.value = 2*pnorm(-abs(z)), 
                       nlog_p = -log10(p.value))
         })) %>%
  unnest(gene_residuals, .drop = TRUE) %>%
  mutate(control = grepl('_', `Gene Symbol`)) # We assume controls have an underscore in their gene name
head(gene_residuals)
```

We can visualize the gene-level residuals for the first treatment condition
```{r}
treatment_data <- gene_residuals %>%
  filter(treatment == random_treatment)
  
ggplot(treatment_data %>% arrange(control) ) +
    aes(x = Avg.Residual, y = nlog_p, label = `Gene Symbol`, 
        color = control) +
    scale_color_manual(values = c('black', 'grey')) +
    geom_point() +
    geom_rug() +
    theme_classic() +
    theme(aspect.ratio = 1, text = element_text(size = 12), 
          plot.title = element_text(size = 12)) +
    ggrepel::geom_label_repel(data = treatment_data %>% top_n(5, nlog_p), 
                     label.padding = 0.1, size = 2.5,  
                     color = 'blue') +
    ggtitle(paste(random_treatment, 'Genes'))
```

# Save Data
```{r}
write_csv(guide_residuals_mapped,here(paste(output_file_root, 'guide_residuals.csv', sep = '_')))
write_csv(gene_residuals,here(paste(output_file_root, 'gene_residuals.csv', sep = '_')))
```

